Variation generation and mean fitness increase: \\ Necessity is the mother of genetic invention

### 2017-11-10 ms.tex
edited OK comments

% Abstract
\begin{abstract}
Generation of variation may be detrimental in well-adapted populations evolving under constant selection.
In a constant environment, genetic modifiers that reduce the rate at which variation is generated by processes such as mutation and migration, succeed.
However, departures from this \emph{reduction principle} have been demonstrated.
Here we analyze a general model of evolution under constant selection where the rate at which variation is generated can depend on the fitness of the individual.
We find that if a modifier allele increases the rate at which individuals of below-average fitness generate variation, then it will increase in frequency and increase the  population mean fitness.
This principle applies to phenomena such as stress-induced mutagenesis, condition-dependent dispersal and outcrossing,
and demonstrates the proverb: \emph{"Necessity is the mother of genetic invention"}.
\end{abstract}

@@

According to the \emph{reduction principle}, in populations that evolve near an equilibrium between natural selection and a process that generates variation (i.e. mutation, migration, or recombination), selection favors neutral modifiers that decrease the rate at which variation is generated.

The \emph{reduction principle} was demonstrated for modifiers of the recombination~\citep{Feldman1972}, mutation, and migration~\citep{Liberman1986a}, and dispersal rates~\citep{Altenberg1987}.
These separate principles were recently unified by~\citet{Altenberg2017}

@@

Another departure from the \emph{linear variation} assumption of the \emph{reduction principle} for mutation rates involves \emph{stress-induced mutagenesis}, a mechanism in which the mutation rate increases in individuals with poor fitness.
Empirical evidence suggests \emph{stress-induced mutagenesis} is common in bacteria and yeast, and may be prevalent in plants, flies, and human cancer cells~\citep{Rosenberg2012,Fitzgerald2017b}.
\citet{Ram2012} demonstrated that modifiers that increase the mutation rate of individuals with below-average fitness actually \textbf{increase} the population mean fitness, rather than decrease it.
Their analysis assumed an infinite population size and constant-viability selection, where the only departure from the \emph{reduction principle} assumptions is due to the differential effect the modifier has on different genotypes.
A similar result has been demonstrated for fitness-associated recombination~\citep{Hadany2003a} and for condition-dependent sexual reproduction~\citep{Hadany2007a}, and evidence suggests that both mechanisms are common in nature~\citep{Ram2016}.

@@

Here, we present a general evolutionary model where the rate at which variation is generated depends on the type of the individual, and specifically on its fitness.
Our results show that the population mean fitness increases if below-average individuals produce more variation than above-average individuals,
and that modifier alleles that induce below-average to produce more variation are favored by natural selection.

@@

\paragraph*{General model.}\label{sec:general_model}
Consider a large haploid population with an arbitrary set of types $A_1, A_2, \ldots, A_n$.
The change in the frequencies of these types is described by the transformation $f \to f'$: 
\begin{equation}\label{eq:model}
\bar{w} f' = (I-C+MC)D f,
\end{equation}
where $f_k$ is the frequency of type $A_k$: $f=(f_1, f_2, \ldots, f_n)$ is a stochastic vector such that $\sum_{k=1}^n{f_k} = 1$;
$C_k$ is the probability that an individual of type $A_k$ will transition to another type ($C$ is a diagonal non-negative matrix with diagonal elements $C_k$);
$M_{i,j}$ is the transition probability from type $A_j$ to type $A_i$ given that an individual of type $A_j$ is transitioning ($M$ is a column-stochastic matrix: $M_{i,j} \ge 0$ for all $i,j$ and $\sum_{j=1}^n {M_{i,j}} = 1$ for all $j$); 
$D_k$ is the relative fitness of type $A_k$ ($D$ is a diagonal matrix with diagonal elements $D_k$, where we can assume $0 < D_k \le 1$, i.e., fitness values are relative to the fitness of the fittest type); 
$I$ is the identity matrix;
and $\bar{w}$ is the normalizing factor such that $\sum_{k=1}^n{f'_k}=1$ and is equal to the population mean fitness (see~\nameref{sec:AppA}).

@@

Transition between types is determined by a combination of two effects:
(i) the probability to transition \emph{out} of type $A_k$ is determined by $C_k$;
(ii) given a transition out of type $A_k$, the distribution of the destination types $A_j$ is given by $M_{j,k}$ (note the index order).
Importantly, different types can have different rates; that is, $C_j \ne C_i$ for some $i,j$. The case $C_j = C_j$ for all $i,j$ is covered by the \emph{reduction principle}~\citep[see][]{Altenberg2017}.

@@

This model can be rephrased in terms of the general model: 
Eq.~\ref{eq:mutation_model_1} is equivalent to Eq.~\ref{eq:model} if we set
\begin{equation} 
C_k = \mu_k, \quad
D_k = w_k, \quad
M_{i,j} = \begin{cases}
\frac{1}{n-1}, & i \ne j \\
0, & i=j
\end{cases}.
\end{equation}

@@

In terms of the general model, we can write
\begin{equation}
C_k = m_k, \quad
D_k = w_k,
%M_{i,j} = \begin{cases}
%\beta, & i = j-1 \\
%1-\beta-\delta, & i = j \\
%\delta, & i = j+1 \\
%1-\delta, & i=j=0 \\
%0, & \text{otherwise}
%\end{cases}
\end{equation}
and
\begin{equation}
M_{i,j} = \begin{bmatrix}
1-\delta & \beta & 0 &  \cdots \\
\delta & 1-\delta-\beta & \beta &  \ddots \\
0 & \delta & 1-\delta-\beta & \ddots \\
\vdots & \ddots & \ddots & \ddots 
\end{bmatrix},
\end{equation}
in which case Eq.~\ref{eq:mutation_model_2} is a specific case of Eq.~\ref{eq:model}.

@@

To rephrase this model in terms of the general model, we define
\begin{equation}
\begin{aligned}
f = (x_1, \ldots, x_n, y_1, \ldots, y_n)\tr, \\
C = diag[m_1, \ldots, m_n, m_1, \ldots, m_n], \\
D = diag[w_1, \ldots, w_n, q_1, \ldots, q_n], \\
M = \begin{bmatrix}
0 & I_n \\
I_n & 0
\end{bmatrix}
\end{aligned}
\end{equation}
where $I_n$ is the n-by-n identity matrix.

@@

Our analysis focuses on populations at equilibrium.
Nevertheless, it has been demonstrated that a modifier which increases the mutation rate of maladapted individuals can be favored by selection~\citep{Ram2012} and reduce the adaptation rate~\citep{Ram2014}.
Similar results have been demonstrated for a modifier that increases the recombination rate in maladapted individuals~\citep{Hadany2003a,Hadany2003b}.

@@

General examples of departures from the \emph{reduction principle} are rare~\citep{Altenberg2017}.
However, it has been demonstrated that higher mutation, recombination, and migration rates are favored by selection in populations evolving under fluctuating selection; see~\citet{Carja2014} and references therein.
Here we have provided another general example, which suggests that a modifier allele that prompts maladapted or below-average individuals increase the rate at which variation is generated, will be favored by selection and will lead to increased population mean fitness.

@@

% Appendix D: inner stability
\subsection*{Appendix D}\label{sec:AppD}
\begin{lemma}[Stability of the equilibrium $v$]
The equilibrium $v$ of the system in Eq.~\ref{eq:model_equilibrium} is locally stable.
In addition,
\begin{enumerate}
\item If $M$ is irreducible and $D$ and $C$ are positive, then the equilibrium is globally stable.
\item If $M$ is irreducible and either $D$ or $C$ are non-negative but not strictly positive, then the equilibrium is globally stable except possibly for initial conditions on the boundary, i.e., when starting with zero elements in the initial frequency vector.
\end{enumerate}
\end{lemma}
\begin{proof}
Set $A=(I-C+MC)D$ and rewrite Eq.~\ref{eq:model} such that $f(t)$ is the frequency vector at time $t$ and
\begin{equation}
f(t) = \frac{Af(t-1)}{\norm{Af(t-1)}},
\end{equation}
where $\norm{f(t)} = \sum_{k=1}^{n}f_k(t)$.
Denote the eigenvalues of $A$ as $\lambda_1, \ldots, \lambda_n$ such that $\abs{\lambda_1} \ge \ldots \ge \abs{\lambda_n}$ and $\lambda_1$ is the \emph{Perron} leading eigenvalue, that is, the largest real eigenvalue with corresponding real non-negative right eigenvector $v$ ($\lambda_1$ is shown in~\nameref{sec:AppA} to be the stable population mean fitness $\bar{w}^*$).

First, note that the normalization $x \goesto \frac{1}{\norm{x}}$ can be applied once instead of $t$ times~\citep[Appendix~A]{Wakano2004}, that is,
\begin{equation}
f(t) = \frac{A^t f(0)}{\norm{A^t f(0)}}.
\label{eq:ftplus1}
\end{equation}

We write $f(0)$ as a linear combination of eigenvectors $v^k$ of $A$ for $k=1, \ldots, n$:
\begin{equation}
f(0) = a_1 v^1 + a_2 v^2 + \ldots + a_n v^n,
\label{eq:linear_combination}
\end{equation}
such that the eigenvector corresponding to the leading eigenvalue is $v=v_1$.

To proceed we need to make sure that $a_1 \ne 0$.
If $M$ is reducible then $A$ is reducible. 
In this case, $a_1 \ne 0$ can be guaranteed by setting $f(0)=v + \epsilon$ such that $\epsilon_k$ is small enough and $v_k+\epsilon_k \ge 0$ for all $k$, and $\sum_{k=1}^{n}{\epsilon_k} = 0$.

If $M$ is irreducible and $D$ and $C$ are positive, then $A$ is irreducible and  any choice of $f(0)$ guarantees $a_1 \ne 0$.

If $M$ is reducible and either $D$ or $C$ are non-negative but not strictly positive, than any choice of $f(0)$ such that $f_k(0)>0$ for all $k$ guarantees $a_1 \ne 0$.

We write, using Eq.~\ref{eq:linear_combination},
\begin{multline}
A^t f(0) = 
a_1 A^t v^1 + a_2 A^t v^2 + \ldots + a_n A^t v^n = \\
a_1 \lambda_1^t v^1 + a_2 \lambda_2^t v^2 + \ldots + a_n \lambda_n^t v^n = \\
\lambda_1^t \bigg(a_ 1v^1 + a_2 \bigg(\frac{\lambda_2}{\lambda_1}\bigg)^t v^2 + \ldots +  a_n \bigg(\frac{\lambda_n}{\lambda_1}\bigg)^t v^n\bigg).
\end{multline}
Therefore,
\begin{equation}
\frac{A^t f(0)}{\lambda_1^t} \goesto a_1 v_1
\end{equation}
as $t \goesto \infty$, because $\abs{\frac{\lambda_k}{\lambda_1}} < 1$
for all $k=2, \ldots, n$.
Finally, 
\begin{equation}
f(t) = 
\frac{A^t f(0)}{\norm{A^t f(0)}} =
\frac{\frac{A^t f(0)}{\lambda_1^t}}{\norm{\frac{A^t f(0)}{\lambda_1^t}}} \goesto
\frac{a_1 v^1}{\norm{a_1 v^1}} = v,
\end{equation}
as $t \goesto \infty$, which completes the proof.
\end{proof}

@@

\bibliographystyle{agsm}
\bibliography{/Users/yoavram/Documents/library}
%\bibliography{ms}

>>>>

% Abstract
\begin{abstract}
Generation of variation may be detrimental in well-adapted populations evolving under constant selection.
In a constant environment, genetic modifiers that reduce the rate at which variation is generated by processes such as mutation and migration, succeed.
However, departures from this \emph{reduction principle} have been demonstrated.
Here we analyze a general model of evolution under constant selection where the rate at which variation is generated can depend on the fitness of the individual.
We find that if a modifier allele increases the rate at which individuals of below-average fitness generate variation, then it will increase in frequency and increase the  population mean fitness.
This principle applies to phenomena such as stress-induced mutagenesis, condition-dependent dispersal and fitness-associated outcrossing,
and exhibits the paraphrase: \emph{"Necessity is the mother of genetic invention."}
\end{abstract}

@@

According to the \emph{reduction principle}, in populations at a balance between natural selection and a process that generates variation (i.e. mutation, migration, or recombination), selection favors neutral modifiers that decrease the rate at which variation is generated. 
The \emph{reduction principle} was demonstrated for modifiers of the recombination~\citep{Feldman1972}, mutation, and migration~\citep{Liberman1986a}, and dispersal rates~\citep{Altenberg1987}.
These separate principles were recently unified by~\citet{Altenberg2017}

@@

Another departure from the \emph{linear variation} assumption of the \emph{reduction principle} for mutation rates involves a mechanism in which the mutation rate increases in individuals with poor fitness -- a mechanism first observed in stressed bacteria~\citep{Foster2007}, although not in a constant environment.
\citet{Ram2012} demonstrated that even in a constant environment, modifiers that increase the mutation rate of individuals with below-average fitness increase the population mean fitness, rather than decrease it.
Their analysis assumed an infinite population size and constant-viability selection in which fitness is determined by the number of mutant alleles accumulated in the genotype.
The only departure from the \emph{reduction principle} assumptions was due to the differential effect the modifier has on different genotypes.
A similar result has been demonstrated for fitness-associated recombination~\citep{Hadany2003a} and for condition-dependent sexual reproduction~\citep{Hadany2007a}, and evidence suggests that both mechanisms are common in nature~\citep{Ram2016}.

@@

Here, we present a general evolutionary model where the the type of the individual determines both its fitness and the rate at which it generates variation.
Our results show that the population mean fitness increases if individuals with below-average fitness produce more variation than individuals with above-average fitness,
and that modifier alleles that induce below-average individuals to produce more variation are favored by natural selection.

@@

\paragraph*{General model.}\label{sec:general_model}
Consider a large haploid population with an arbitrary set of types $A_1, A_2, \ldots, A_n$.
The frequency and fitness of individuals of type $A_k$ are $f_k$ and $D_k$, respectively.
The probability that an individual of type $A_k$ will transition to some other type is $C_k$, and given a transition occurs, the probability that it will transition to type $A_i$ is $M_{i,k}$.
Therefore, the change in the frequencies of types $A_k$ is described by the transformation $f \to f'$: 
\begin{equation}\label{eq:model}
\bar{w} f' = (I-C+MC)D f,
\end{equation}
where $f=(f_1, f_2, \ldots, f_n)$ is a stochastic vector such that $\sum_{k=1}^n{f_k} = 1$;
$D$ is a diagonal matrix with diagonal elements $D_k$, where we can assume $0 < D_k \le 1$, i.e., fitness values are relative to the fitness of the fittest type; 
$C$ is a diagonal non-negative matrix with diagonal elements $C_k$;
$M$ is a column-stochastic matrix: $M_{i,k} \ge 0$ for all $i,k$ and $\sum_{k=1}^n {M_{i,k}} = 1$ for all $j$; 
$I$ is the identity matrix;
and $\bar{w}$ is the normalizing factor such that $\sum_{k=1}^n{f'_k}=1$ and is equal to the population mean fitness (see~\nameref{sec:AppA}).

@@

Transition between types is determined by a combination of two effects:
(i) the probability to transition \emph{out} of type $A_k$ is determined by $C_k$;
(ii) given a transition out of type $A_k$, the distribution of the destination types $A_i$ is given by $M_{i,k}$ (note the index order).
Importantly, different types can have different rates; that is, $C_i \ne C_j$ for some $i,j$. The case $C_i = C_j$ for all $i,j$ is covered by the \emph{reduction principle}~\citep[see][]{Altenberg2017}.

@@

This model can be rephrased in terms of the general model: 
Eq.~\ref{eq:mutation_model_1} is equivalent to Eq.~\ref{eq:model} if we set
\begin{equation} 
C_k = \mu_k, \quad
D_k = w_k, \quad
M_{i,j} = \begin{cases}
\frac{1}{n-1}, & i \ne j \\
0, & i=j
\end{cases}.
\end{equation}
Note that here $M$ is irreducible and therefore so is $(I-C+MC)D$.

@@

In terms of the general model, we can write
\begin{equation}
C_k = m_k, \quad
D_k = w_k,
%M_{i,j} = \begin{cases}
%\beta, & i = j-1 \\
%1-\beta-\delta, & i = j \\
%\delta, & i = j+1 \\
%1-\delta, & i=j=0 \\
%0, & \text{otherwise}
%\end{cases}
\end{equation}
and
\begin{equation}
M_{i,j} = \begin{bmatrix}
1-\delta & \beta & 0 &  \cdots \\
\delta & 1-\delta-\beta & \beta &  \ddots \\
0 & \delta & 1-\delta-\beta & \ddots \\
\vdots & \ddots & \ddots & \ddots 
\end{bmatrix}.
\end{equation}
Here, too, $M$ and $(I-C+MC)D$ are irreducible.

@@

To rephrase this model in terms of the general model, we define
\begin{equation}
\begin{aligned}
f = (x_1, \ldots, x_n, y_1, \ldots, y_n)\tr, \\
C = diag[m_1, \ldots, m_n, m_1, \ldots, m_n], \\
D = diag[w_1, \ldots, w_n, q_1, \ldots, q_n], \\
M = \begin{bmatrix}
0 & I_n \\
I_n & 0
\end{bmatrix}
\end{aligned}
\end{equation}
where $I_n$ is the n-by-n identity matrix.
In this case, $M$ is reducible and so is $(I-C+MC)D$.

@@

Our analysis focuses on populations at equilibrium.
Nevertheless, it has been demonstrated that a modifier which increases the mutation rate of maladapted individuals can be favored by selection~\citep{Ram2012} and reduce the adaptation rate~\citep{Ram2014},
and empirical evidence suggests that \emph{stress-induced mutagenesis} is common in bacteria and yeast, and may be prevalent in plants, flies, and human cancer cells~\citep{Rosenberg2012,Fitzgerald2017b}.
Similar theoretical results have been demonstrated for a modifier that increases the recombination rate in maladapted individuals~\citep{Hadany2003a,Hadany2003b}.

@@

General examples of departures from the \emph{reduction principle} are rare~\citep{Altenberg2017}.
For example, it has been demonstrated that higher mutation, recombination, and migration rates are favored by selection in populations evolving under fluctuating selection; see~\citet{Carja2014} and references therein.
Here we have provided another general example, which suggests that a modifier allele that prompts individuals with below-average fitness to increase the rate at which variation is generated, will be favored by selection and will lead to increased population mean fitness.

@@

% Appendix D: inner stability
\subsection*{Appendix D}\label{sec:AppD}
\begin{lemma}[Stability of the equilibrium $v$]
The equilibrium $v$ of the system in Eq.~\ref{eq:model_equilibrium} is locally stable.
\end{lemma}
\begin{proof}
Set $A=(I-C+MC)D$ and rewrite Eq.~\ref{eq:model} such that $f(t)$ is the frequency vector at time $t$ and
\begin{equation}
f(t) = \frac{Af(t-1)}{\norm{Af(t-1)}},
\end{equation}
where $\norm{f(t)} = \sum_{k=1}^{n}f_k(t)$.
Denote the eigenvalues of $A$ as $\lambda_1, \ldots, \lambda_n$ such that $\abs{\lambda_1} \ge \ldots \ge \abs{\lambda_n}$ and $\lambda_1$ is the \emph{Perron} leading eigenvalue, that is, the largest real eigenvalue with corresponding real non-negative right eigenvector $v$ ($\lambda_1$ is shown in~\nameref{sec:AppA} to be the stable population mean fitness $\bar{w}^*$).

First, note that the normalization $x \goesto \frac{1}{\norm{x}}$ can be applied once instead of $t$ times~\citep[Appendix~A]{Wakano2004}, that is,
\begin{equation}
f(t+k) = \frac{A^t f(k)}{\norm{A^t f(k)}}.
\label{eq:ftplus1}
\end{equation}

To check local stability, we make a perturbation $\epsilon$ to $v$, such that this perturbation is small enough so that when $v+\epsilon$ is written as a linear combination of $\{v^k\}_{k=1}^{n}$ the eigenvalues of $A$, the coefficient of $v=v^1$ is not zero.
That is, the frequency vector after the perturbation $f(0)=v+\epsilon$ can be written
\begin{equation}
f(0) = a_1 v^1 + \ldots + a_n v^n,
\label{eq:linear_combination}
\end{equation}
with $\sum_{k=1}^{n}{f_k(0)}=1$ and $a_1 \ne 0$.

We now check that $f(t) \goesto v$ as $t \goesto \infty$, which will indicate that $v$ is locally stable.
We write, using Eq.~\ref{eq:linear_combination},
\begin{multline}
A^t f(0) = 
a_1 A^t v^1 + a_2 A^t v^2 + \ldots + a_n A^t v^n = \\
a_1 \lambda_1^t v^1 + a_2 \lambda_2^t v^2 + \ldots + a_n \lambda_n^t v^n = \\
\lambda_1^t \bigg(a_ 1v^1 + a_2 \bigg(\frac{\lambda_2}{\lambda_1}\bigg)^t v^2 + \ldots +  a_n \bigg(\frac{\lambda_n}{\lambda_1}\bigg)^t v^n\bigg).
\end{multline}
Therefore,
\begin{equation}
\frac{A^t f(0)}{\lambda_1^t} \goesto a_1 v_1
\end{equation}
as $t \goesto \infty$, because $\abs{\frac{\lambda_k}{\lambda_1}} < 1$
for all $k=2, \ldots, n$.
Finally, 
\begin{equation}
f(t) = 
\frac{A^t f(0)}{\norm{A^t f(0)}} =
\frac{\frac{A^t f(0)}{\lambda_1^t}}{\norm{\frac{A^t f(0)}{\lambda_1^t}}} \goesto
\frac{a_1 v^1}{\norm{a_1 v^1}} = v,
\end{equation}
as $t \goesto \infty$, which completes the proof.
\end{proof}

@@

\bibliographystyle{agsm}
%\bibliography{/Users/yoavram/Documents/library}
\bibliography{ms}
